BOLD Parameter Estimation Using Particle Filters

### 2010-09-02 doc/thesis/results.tex
finished I think

The slice simulation was performed using
Physics-Oriented Simulated Scanner for Understanding MRI (POSSUM) which
models noise realistically. POSSUM demonstrates the
particle filter modeling the \ac{BOLD} signal on a large scale.
In the POSSUM simulation there were four
discrete parameter sets driving the time series, although each voxel had a
different tissue composition and a different
noise realization. Therefore the POSSUM simulation was
a good test of the applicability of particle filters for performing
whole brain analysis.

@@

\begin{figure}
\centering
\includegraphics[clip=true,trim=1cm 0cm 0cm 0cm, width=17cm]{images/long_converge}
\caption[\ac{BOLD} estimate converging for a very long \ac{fMRI} run.]{\ac{BOLD} estimate 
converging for a 2400 second \ac{fMRI} run. Darker bars indicate
bins with more particles with BOLD estimate in that range. ($\sigma_x = 0.0005$, $\sigma_y = 0.001$)}
\label{fig:long_converge}
\end{figure}

\begin{figure}
\centering
\includegraphics[clip=true,trim=1cm 0cm 0cm 0cm, width=17cm]{images/long_converge_500}
\caption[First 500 seconds of the \ac{BOLD} estimate converging for a very 
long \ac{fMRI} run.]{First 500 seconds of \ac{BOLD} estimate 
converging for a 2400 second \ac{fMRI} run. Darker bars indicate
bins with more particles with BOLD estimate in that range. ($\sigma_x = 0.0005$, $\sigma_y = 0.001$)}
\label{fig:long_converge_500}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[clip=true,trim=6cm 3cm 6cm 3cm,height=9cm]{images/fits_noiseonly}
\caption[Results for non-active, low noise signal]  ($\sigma_x = 0.005$, $\sigma_y = 0.01$)
{Comparison of the estimated BOLD signal for the low noise signal-free case.
 ($\sigma_x = 0.005$, $\sigma_y = 0.01$). Note the line thickness is caused
by all the estimates overlapping.}
\label{fig:fits_noiseonly}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[clip=true,trim=6cm 3cm 6cm 3cm,height=9cm]{images/justnoise_fit_0}
\caption{Single Fit Results for non-active, low noise signal.}
\label{fig:justnoise_fit_0}
\end{figure} %uses allnoise/ALLNOISE-0-w0

@@

\begin{figure}[H]
\centering
\includegraphics[clip=true,trim=6cm 3cm 6cm 3cm,height=9cm]{images/justbignoise_fit_0}
\caption{Single Fit Results for non-active, high noise signal.}
\label{fig:justbignoise_fit_0}
\end{figure} %uses allnoise/ALLNOISE-0-w0

@@

\subsection{Single Voxel Summary}
\label{sec:SingleVoxelReview}
Because of the variability in the signal levels, the
raw \ac{RMSR} cannot be used to rate the fit. As demonstrated by \autoref{tab:NoiseOnlyResults},
a low \ac{RMSR} does not necessarily indicate a good fit.
Therefore, a normalized version of the \ac{RMSR} was used. To normalize the
residual, both the estimated and preprocessed \ac{BOLD} signals were divided by an estimator
of scale and then the new \ac{RMSR} was calculated.
Considering the tendency of \ac{fMRI} to have large unexplainable
peaks and troughs, the Median-Absolute-Deviation (MAD) was used (\autoref{eq:mad}).
This is an estimator of the standard
deviation, and thus a good estimator of the scale of the input signal.
The normalized \ac{RMSR} values are
shown in \autoref{tab:SingleVoxelActivationComparison}. A second potential method of
gauging performance is mutual information. Mutual information is a method of measuring
the interdependence of two random variables. If two signals are truly independent,
then the mutual information will be zero. Although ideally suited to discrete
distributions, by using histograms
it is possible to derive a joint distribution of two signals. The algorithm for
mutual information is based on that joint distribution:
\begin{equation}
\sum_{x,y} p(x,y) \log_2\left(\frac{p(x,y)}{p(x)p(y)}\right)
\end{equation}
Unfortunately the number of bins causes bias in the output, thus to correct for this,
I subtracted the estimated bias:
\begin{equation}
\text{bias} = \frac{N_{bins}}{2Nlog(2)}
\end{equation}
where $N$ is the number of samples and $N_{bins}$ is the number of bins. For all
the mutual information estimates in this work 6 bins were used for the marginal
distribution of each signal. Additionally, throughout log base 2 will be used.
This leads to 36 total bins in the joint, so the bias is:
\begin{equation}
\text{bias} = \frac{18}{N}
\end{equation}
 Note that subtracting the bias can result in negative mutual
information, which should not technically be possible; so any negative mutual information
was taken as 0.

@@

\begin{figure}
\centering
\subfigure[Region labels for simulated slice.]{
\includegraphics[height=8cm]{images/simregions.png} \label{fig:simslice_mask} }
\subfigure[Normalized \ac{RMSR} Map. Scale is Normalized \ac{RMSE}. Lower (yellow) is better.]
{\label{fig:simslice_hm_res} \includegraphics[height=8cm]{images/sim_hm}}
\subfigure[Mutual Information Map. Scale is bits. Higher (yellow) is better.]
{\label{fig:sim_hm_mi} \includegraphics[height=8cm]{images/sim_hm_mi}}
\subfigure[SNR Map of POSSUM simulated data. Mean SNR per region, Region 1: $0.8$, Region 2: $0.97$, and
Region 3: $0.39$.]{\includegraphics[height=8cm]{images/snr_hm.png}\label{fig:SimSNRhm}}
\caption{Comparison of activation with greymatter, parameter regions.}
\label{fig:simslice_hm}
\end{figure}

To test the usefulness of the particle filter on a larger scale, a modified version
of the FSL tool
POSSUM was used to generate an entire \ac{fMRI} image from a parameter map. The parameter map was generated
by taking an existing activation map and assigning discrete parameter sets to each region.
The result was a four dimensional (length x width
x height x parameter) image with spatially varying parameters. Possum was then modified
to take a parameter map and generate activation levels depending on the parameters at that
point. The patch for POSSUM will be made available. Note that the noise
level was set to an SNR of 20, but due to changes
in the program the true signal-to-noise ratio was much lower, as seen in the
in \autoref{fig:SimSNRhm}. The mean SNR for each region was calculated only for the
voxels with a Signal-To-Noise ratio above $0.1$.

@@

\begin{figure} %bottom left
\centering
\includegraphics[clip=true,trim=2.5cm 2cm 2cm 1cm,width=15cm]{images/slicesim_hist1}
\caption{Histogram of estimated parameters in section 1 in voxels 
with mutual information greater than $0.15$} \label{fig:slicesim_hist1}
\end{figure}

\begin{figure} %top left
\centering
\includegraphics[clip=true,trim=2.5cm 2cm 2cm 1cm,width=15cm]{images/slicesim_hist2}
\caption{Histogram of estimated parameters in section 2 in voxels with mutual information greater
than $0.15$}
\label{fig:slicesim_hist2}
\end{figure}

\begin{figure} %top right
\centering
\includegraphics[clip=true,trim=2.5cm 2cm 2cm 1cm,width=15cm]{images/slicesim_hist3}
\caption{Histogram of estimated parameters in section 3 in voxels with mutual information greater
than $0.15$}
\label{fig:slicesim_hist3}
\end{figure}

@@

The histograms again demonstrate that a single point estimate of the parameters
is elusive for this set of parameters. However the data clearly show the power
of the particle filter at identifying regions of activation.
The thresholds applied to this slice, both for mutual information and
\acp{RMSR} are arbitrary. Applying a
threshold is helpful for visualization but is rarely useful for further analysis.
As noted in \autoref{sec:SingleVoxelReview},
the false positives present in the Mutual Information
map are different from those in the \ac{RMSR} map. This furthers the argument
for combining the two metrics to increase power. Although
at first glance it would appear that there are false negatives in
\autoref{fig:sim_hm_mi}; this is
not actually the case. POSSUM simulates different tissues, and white matter
does not typically  have a \ac{BOLD} response. This explains some of the 
holes in regions
2 and 3. These results certainly indicate that the particle filter is effective
at regressing against a noisy signal.

>>>>

The slice simulation was performed using
\ac{POSSUM} which
models noise realistically. \ac{POSSUM} demonstrates the
particle filter modeling the \ac{BOLD} signal on a large scale.
In the \ac{POSSUM} simulation there were four
discrete parameter sets driving the time series, although each voxel had a
different tissue composition and a different
noise realization. Therefore the \ac{POSSUM} simulation was
a good test of the applicability of particle filters for performing
whole brain analysis.

@@

\begin{figure}
\centering
\includegraphics[clip=true,trim=1cm 0cm 0cm 0cm, width=17cm]{images/long_converge}
\caption[\ac{BOLD} estimate converging for a very long \ac{fMRI} run.]{\ac{BOLD} estimate 
converging for a 2400 second \ac{fMRI} run. Darker bars indicate
bins with more particles with \ac{BOLD} estimate in that range. ($\sigma_x = 0.0005$, $\sigma_y = 0.001$)}
\label{fig:long_converge}
\end{figure}

\begin{figure}
\centering
\includegraphics[clip=true,trim=1cm 0cm 0cm 0cm, width=17cm]{images/long_converge_500}
\caption[First 500 seconds of the \ac{BOLD} estimate converging for a very 
long \ac{fMRI} run.]{First 500 seconds of \ac{BOLD} estimate 
converging for a 2400 second \ac{fMRI} run. Darker bars indicate
bins with more particles with \ac{BOLD} estimate in that range. ($\sigma_x = 0.0005$, $\sigma_y = 0.001$)}
\label{fig:long_converge_500}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[clip=true,trim=6cm 3cm 6cm 3cm,height=9cm]{images/fits_noiseonly}
\caption[Results for non-active, low noise signal]  
{Comparison of the estimated \ac{BOLD} signal for the low noise signal-free case.
 ($\sigma_x = 0.005$, $\sigma_y = 0.01$). Note the line thickness is caused
by all the estimates overlapping.}
\label{fig:fits_noiseonly}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[clip=true,trim=6cm 3cm 6cm 3cm,height=9cm]{images/justnoise_fit_0}
\caption[Single Fit Results for non-active, low noise signal.]
{Single Fit Results for non-active, low noise signal
 ($\sigma_x = 0.005$, $\sigma_y = 0.01$).}
\label{fig:justnoise_fit_0}
\end{figure} %uses allnoise/ALLNOISE-0-w0

@@

\begin{figure}[H]
\centering
\includegraphics[clip=true,trim=6cm 3cm 6cm 3cm,height=9cm]{images/justbignoise_fit_0}
\caption[Single Fit Results for non-active, high noise signal.]
{Single Fit Results for non-active, high noise signal
 ($\sigma_x = 0.005$, $\sigma_y = 0.01$).}
\label{fig:justbignoise_fit_0}
\end{figure} %uses allnoise/ALLNOISE-0-w0

@@

\subsection{Single Voxel Summary}
\label{sec:SingleVoxelReview}
Because of the variability in the signal levels, the
raw \ac{RMSR} cannot be used to rate the fit. As demonstrated by \autoref{tab:NoiseOnlyResults},
a low \ac{RMSR} does not necessarily indicate a good fit.
Therefore, a normalized version of the \ac{RMSR} was used. To normalize the
residual, both the estimated and preprocessed \ac{BOLD} signals were divided by an estimator
of scale and then the new \ac{RMSR} was calculated.
Considering the tendency of \ac{fMRI} to have large unexplainable
peaks and troughs, the \ac{MAD} was used (\autoref{eq:mad}).
This is an estimator of the standard
deviation, and thus a good estimator of the scale of the input signal.
The normalized \ac{RMSR} values are
shown in \autoref{tab:SingleVoxelActivationComparison}. A second potential method of
gauging performance is mutual information. Mutual information is a method of measuring
the interdependence of two random variables. If two signals are truly independent,
then the mutual information will be zero. Although ideally suited to discrete
distributions, by using histograms
it is possible to derive a joint distribution of two signals. The algorithm for
mutual information is based on that joint distribution:
\begin{equation}
\sum_{x,y} p(x,y) \log_2\left(\frac{p(x,y)}{p(x)p(y)}\right)
\end{equation}
Unfortunately the number of bins causes bias in the output, thus to correct for this,
I subtracted the estimated bias:
\begin{equation}
\text{bias} = \frac{N_{bins}}{2Nlog(2)}
\end{equation}
where $N$ is the number of samples and $N_{bins}$ is the number of bins. For all
the mutual information estimates in this work 6 bins were used for the marginal
distribution of each signal. Additionally, throughout log base 2 will be used.
This leads to 36 total bins in the joint, so the bias is:
\begin{equation}
\text{bias} = \frac{18}{N}
\end{equation}
 Note that subtracting the bias can result in negative mutual
information, which should not technically be possible; so any negative mutual information
was taken as 0.

@@

\begin{figure}
\centering
\subfigure[Region labels for simulated slice.]{
\includegraphics[height=9cm]{images/simregions.png} \label{fig:simslice_mask} }
\subfigure[\ac{SNR} Map of \ac{POSSUM} simulated data. Mean \ac{SNR} per region, 
Region 1: $0.8$, Region 2: $0.97$, and
Region 3: $0.39$.]{\includegraphics[height=9cm]{images/snr_hm.png}\label{fig:SimSNRhm}}

\subfigure[Normalized \ac{RMSR} Map. Scale is Normalized \ac{RMSE}. Lower (yellow) is better.]
{\label{fig:simslice_hm_res} \includegraphics[height=8.5cm]{images/sim_hm}}
\subfigure[Mutual Information Map. Scale is bits. Higher (yellow) indicates better fit.]
{\label{fig:sim_hm_mi} \includegraphics[height=8.5cm]{images/sim_hm_mi}}

\caption{Comparison of activation regions, a) grey matter and regions, b) \ac{SNR} of active
regions, c) \acl{RMSR} Map, d) Mutual Information Map.}
\label{fig:simslice_hm}
\end{figure}

To test the usefulness of the particle filter on a larger scale, a modified version
of the \ac{FSL} tool
\ac{POSSUM} was used to generate an entire \ac{fMRI} image from a parameter map. The parameter map was generated
by taking an existing activation map and assigning discrete parameter sets to each region.
The result was a four dimensional (length x width
x height x parameter) image with spatially varying parameters. Possum was then modified
to take a parameter map and generate activation levels depending on the parameters at that
point. The patch for \ac{POSSUM} will be made available. Note that the noise
level was set to an \ac{SNR} of 20, but due to changes
in the program the true signal-to-noise ratio was much lower, as seen in the
in \autoref{fig:SimSNRhm}. The mean \ac{SNR} for each region was calculated only for the
voxels with a Signal-To-Noise ratio above $0.1$.

@@

\begin{figure} %bottom left
\centering
\includegraphics[clip=true,trim=2.5cm 2cm 2cm 1cm,width=\textwidth]{images/slicesim_hist1}
\caption[Histogram of estimated parameters in section 1]{Histogram of estimated parameters in 
section 1 in voxels 
with mutual information greater than $0.15$} \label{fig:slicesim_hist1}
\end{figure}

\begin{figure} %top left
\centering
\includegraphics[clip=true,trim=2.5cm 2cm 2cm 1cm,width=\textwidth]{images/slicesim_hist2}
\caption[Histogram of estimated parameters in section 2]
{Histogram of estimated parameters in section 2 in voxels with mutual information greater
than $0.15$}
\label{fig:slicesim_hist2}
\end{figure}

\begin{figure} %top right
\centering
\includegraphics[clip=true,trim=2.5cm 2cm 2cm 1cm,width=\textwidth]{images/slicesim_hist3}
\caption[Histogram of estimated parameters in section 3]
{Histogram of estimated parameters in section 3 in voxels with mutual information greater
than $0.15$}
\label{fig:slicesim_hist3}
\end{figure}

@@

The histograms again demonstrate that a single point estimate of the parameters
is elusive for this set of parameters. However the data clearly show the power
of the particle filter at identifying regions of activation.
The thresholds applied to this slice, both for mutual information and
\acp{RMSR} are arbitrary. Applying a
threshold is helpful for visualization but is rarely useful for further analysis.
As noted in \autoref{sec:SingleVoxelReview},
the false positives present in the Mutual Information
map are different from those in the \ac{RMSR} map. This furthers the argument
for combining the two metrics to increase power. Although
at first glance it would appear that there are false negatives in
\autoref{fig:sim_hm_mi}; this is
not actually the case. \ac{POSSUM} simulates different tissues, and white matter
does not typically  have a \ac{BOLD} response. This explains some of the 
holes in regions
2 and 3. These results certainly indicate that the particle filter is effective
at regressing against a noisy signal.
